Analysis for Biracial Grade Retention Project

### 2021-11-24 analysis/analysis.Rmd
Change caption to fig.cap

```{r check-retention-code, caption="The proportion of students measured at below grade level by age and current grade. This graph is used to ensure that the dependent variable is calculated correctly."}
acs %>%
  group_by(age, current_grade) %>%
  summarize(below=mean(below_exp_grade)) %>%
  ggplot(aes(x=current_grade, y=age, fill=below))+
  geom_tile(color="grey20")+
  geom_text(aes(label=below))+
  theme_bw()+
  scale_fill_viridis_c()+
  scale_y_continuous(breaks=5:20, minor_breaks = NULL)+
  labs(x="current grade", y="age", fill="proportion below grade level")
```

@@

```{r retention-by-grade, caption="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
acs %>%
  group_by(current_grade) %>%
  summarize(below_grade=mean(below_exp_grade),
            n=n(),
            se=sqrt((below_grade*(1-below_grade))/n)) %>%
  ggplot(aes(x=current_grade, y=below_grade, group=1,
             ymin=below_grade-1.96*se, ymax=below_grade+1.96*se))+
  geom_smooth(se=FALSE)+
  geom_line(linetype=2, color="grey30")+
  geom_point()+
  geom_linerange()+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r retention-by-year, caption="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
acs %>%
  group_by(year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade))+
  geom_line()+
  geom_point()+
  geom_smooth(se=FALSE)+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r retention-by-year-grade, caption="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  #geom_line()+
  geom_smooth(se=FALSE)+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r example-plot, caption="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
g <- coef_table_mod0 %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

@@

```{r cond-means-model4, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
coef_table_mod4 <- calculate_marg_means(model4)

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
coef_table_mod1 <- calculate_marg_means(model1)
coef_table_mod2 <- calculate_marg_means(model2)

@@

```{r models-by-grade, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Panels compare results when data are restricted to certain grade ranges."}
#run separate models by grade groups
model4_elem <- update(model4, data=subset(acs, current_grade<="5th"))
model4_middle <- update(model4, data=subset(acs, current_grade>"5th" & 
                                        current_grade<"9th"))
model4_hs <- update(model4, data=subset(acs, current_grade>="9th"))

@@

```{r models-parent-link, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
#run separate models by grade groups
model4_best_link <- update(model4, data=subset(acs, momrule==11 & poprule==11))

@@

```{r cult-resource-model, warning=FALSE, caption="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Cultural resources are determined by a summated score of standardized native-born status, parents' native-born status, English proficieny, and parents' English proficiency. Parents' native-born status is reverse coded based on model results which show it increases grade retention risk."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_cultr <- glm(cult_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    +degree_mother+degree_father+sqrt(family_income/10000)+
                     own_home+parents_married, 
                  data=acs, family="gaussian")
coef_table_cultr <- calculate_marg_means(model_cultr)

@@

```{r fam-resource-model, warning=FALSE, caption="Average family resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Family resources are determined by a summated score of standardized family income, parental education, and home ownership."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_famr <- glm(fam_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    foreign_born+foreign_born_mother+foreign_born_father+
                    spk_eng+spk_eng_mother+spk_eng_father, 
                  data=acs, family="gaussian")
coef_table_famr <- calculate_marg_means(model_famr)

>>>>

```{r check-retention-code, fig.cap="The proportion of students measured at below grade level by age and current grade. This graph is used to ensure that the dependent variable is calculated correctly."}
acs %>%
  group_by(age, current_grade) %>%
  summarize(below=mean(below_exp_grade)) %>%
  ggplot(aes(x=current_grade, y=age, fill=below))+
  geom_tile(color="grey20")+
  geom_text(aes(label=below))+
  theme_bw()+
  scale_fill_viridis_c()+
  scale_y_continuous(breaks=5:20, minor_breaks = NULL)+
  labs(x="current grade", y="age", fill="proportion below grade level")
```

@@

```{r retention-by-grade, fig.cap="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
acs %>%
  group_by(current_grade) %>%
  summarize(below_grade=mean(below_exp_grade),
            n=n(),
            se=sqrt((below_grade*(1-below_grade))/n)) %>%
  ggplot(aes(x=current_grade, y=below_grade, group=1,
             ymin=below_grade-1.96*se, ymax=below_grade+1.96*se))+
  geom_smooth(se=FALSE)+
  geom_line(linetype=2, color="grey30")+
  geom_point()+
  geom_linerange()+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r retention-by-year, fig.cap="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
acs %>%
  group_by(year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade))+
  geom_line()+
  geom_point()+
  geom_smooth(se=FALSE)+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r retention-by-year-grade, fig.cap="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  #geom_line()+
  geom_smooth(se=FALSE)+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
g <- coef_table_mod0 %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

@@

```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
coef_table_mod4 <- calculate_marg_means(model4)

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
coef_table_mod1 <- calculate_marg_means(model1)
coef_table_mod2 <- calculate_marg_means(model2)

@@

```{r models-by-grade, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Panels compare results when data are restricted to certain grade ranges."}
#run separate models by grade groups
model4_elem <- update(model4, data=subset(acs, current_grade<="5th"))
model4_middle <- update(model4, data=subset(acs, current_grade>"5th" & 
                                        current_grade<"9th"))
model4_hs <- update(model4, data=subset(acs, current_grade>="9th"))

@@

```{r models-parent-link, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
#run separate models by grade groups
model4_best_link <- update(model4, data=subset(acs, momrule==11 & poprule==11))

@@

```{r cult-resource-model, warning=FALSE, fig.cap="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Cultural resources are determined by a summated score of standardized native-born status, parents' native-born status, English proficieny, and parents' English proficiency. Parents' native-born status is reverse coded based on model results which show it increases grade retention risk."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_cultr <- glm(cult_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    +degree_mother+degree_father+sqrt(family_income/10000)+
                     own_home+parents_married, 
                  data=acs, family="gaussian")
coef_table_cultr <- calculate_marg_means(model_cultr)

@@

```{r fam-resource-model, warning=FALSE, fig.cap="Average family resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Family resources are determined by a summated score of standardized family income, parental education, and home ownership."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_famr <- glm(fam_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    foreign_born+foreign_born_mother+foreign_born_father+
                    spk_eng+spk_eng_mother+spk_eng_father, 
                  data=acs, family="gaussian")
coef_table_famr <- calculate_marg_means(model_famr)
